Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-20 doc/mc-paper.tex
Tweaked intro

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler
generates a d-DNNF representation, and the {\em proof
  generator} uses this to generate a CRAT file.
The {\em proof checker} verifies the equivalence of the CNF and CRAT representations.
The {\em ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently (and, in some cases, formally) checked
\cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress
has also been made on extending these proof systems to other domains,
including quantified Boolean
formulas~\cite{bryant:cade:2021,heule:JAR2014}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound.  They also present
experimental results for cases where they modify existing model
counters to generate proof traces, and where they generate proof
traces from d-DNNF graphs.
%% They have not formally verified their
%%proof framework or created a formally verified checker.

Our toolchain is not directly comparable to theirs---they directly
certify the numeric value produced by a model counter, while we
certify that the graph generated by a knowledge compiler is logically
equivalent to the input formula.  We can also validate the standard or
weighted count generated from that representation.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our toolchain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

Verified checkers for unsatisfiability include \cite{Cruz:Filipe:et:al:16,Heule:et:al:17,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
In contrast, our proofs construct a new propositional formula, and we have to verify that it
is partitioned and is equivalent to the input.
This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
Sections~\ref{sect:crat} and \ref{section:formally-verified-toolchain}.

>>>>

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a d-DNNF representation, and the {\em proof
  generator} uses this to generate a CRAT file.
The {\em proof checker} verifies the equivalence of the CNF and CRAT representations.
The {\em ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{RAT,wetzler14_drattrim}.
A number formally verified checkers for this task have been developed within different frameworks~\cite{cruz-cade-2017,Heule:et:al:17,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
By contrast, our proofs construct a new propositional formula, and we must verify both that it
is partitioned and that it is equivalent to the input formula.
This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
Sections~\ref{sect:crat} and \ref{section:formally-verified-toolchain}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithms commonly used by model counters.
They present a series of lemmas and a theorem stating
that their checking framework is sound.  They also present
experimental results for cases where they modify existing model
counters to generate proof traces, and where they generate proof
traces from d-DNNF graphs.
%% They have not formally verified their
%%proof framework or created a formally verified checker.
Our toolchain is not directly comparable to theirs---they directly
certify the numeric value produced by a model counter, while we
certify that the graph generated by a knowledge compiler is logically
equivalent to the input formula.  We can then validate a standard or
weighted count from that representation.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our toolchain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.
